---
title: '2023-12 Boxplots for differential expression'
subtitle: RE project
output:
  html_document:
    df_print: paged
    number_sections: false
    toc: yes
    toc_depth: '3'
    toc_float: yes
    code_folding: hide
    theme: cosmo
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(survival)
library(survminer)
library(data.table)
library(table1)
library(reshape2)
library(doParallel)
library(foreach)
library(DESeq2)
library(limma)
library(edgeR)
library(kableExtra)
source("config.R")



```

## Data Cleanup {.tabset}

### 1. Normalizing counts

```{r normalization}
raw_counts <- readRDS(paste0(data_path,'/Data/RE_all_repFamily_1_raw_counts.RDS'))
normal_counts <- calcNormFactors(raw_counts, method = 'RLE')
normal_counts <- as.data.frame(cpm(normal_counts, normalized.lib.sizes=T))
```

### 2. Parsing through IDs

```{r combine-ids}
glioma_dt <- read.csv(paste0(data_path,"/Data/glioma_dt.csv"))
glioma_dt <- data.table(glioma_dt)
glioma_dt$case_submitter_id <- substr(glioma_dt$case_submitter_id, 6, 12)

```

### 3. Recoding gliomas


```{r glioma_reclass}

#GBMS
glioma_dt$WHO2021_recode <- ifelse(glioma_dt$IDH =='WT', 'Glioblastoma, IDH-wt', 'Diffuse Glioma, IDH-mutant, NEC')

#Oligos

glioma_dt <- glioma_dt %>%
  mutate(WHO2021_recode = ifelse(!is.na(x1p19q) & !is.na(IDH) & IDH == 'Mutant' & x1p19q == 'codel',
                                 'Oligodendroglioma, IDH-mutant', WHO2021_recode))

glioma_dt <- glioma_dt %>%
  mutate(WHO2021_recode = ifelse(!is.na(IDH) & IDH == 'Mutant' & TERT == 'Mutant' & WHO2021_recode == 'Diffuse Glioma, IDH-mutant, NEC',
                                 'Oligodendroglioma, IDH-mutant', WHO2021_recode))


#Astros
glioma_dt <- glioma_dt %>%
  mutate(WHO2021_recode = ifelse(!is.na(IDH) & !is.na(ATRX.status) & IDH == 'Mutant' & ATRX.status == 'Mutant',
                                 'Astrocytoma, IDH-mutant', WHO2021_recode))
```



```{r}

count_ids <- c(colnames(normal_counts))
count_ids <- substr(count_ids, 1, 7)

glioma_ids <- glioma_dt$case_submitter_id

glioma_counts <- normal_counts[, grepl(paste(glioma_ids, collapse ="|"), names(normal_counts))]
glioma_counts <- t(glioma_counts)
glioma_counts <- data.frame(glioma_counts)
glioma_counts$IID <- rownames(glioma_counts)
glioma_counts$IID <- rownames(glioma_counts)
rownames(glioma_counts) <- NULL
glioma_counts$IID <- substr(glioma_counts$IID, 1, 7)
rm(list =c('normal_counts', 'raw_counts'))
```



```{r}

glioma_surv <- merge(glioma_counts, glioma_dt, by.x="IID", by.y="case_submitter_id")
glioma_surv$who_grade <- glioma_surv$Grade

glioma_surv$who_grade <- ifelse(glioma_surv$WHO2021_recode == "Glioblastoma, IDH-wt", "G4", glioma_surv$who_grade)

glioma_surv$WHO2021_integrated <- paste(glioma_surv$WHO2021_recode, glioma_surv$who_grade, sep=", ")

num_columns <- sapply(glioma_counts, is.numeric)

```

### Perecentile function

```{r percentile-function}
is_in_top_percentile <- function(x, p) {
  top_percentile <- quantile(x, p, na.rm = TRUE)
  ifelse(x >= top_percentile, 1, 0)
}


```

### Geno's tertile function

```{r}
# Function to group into low mid and high groups

percent_groups <- function(x, p_lower, p_upper) {
  
  low_thresh <- quantile(x, p_lower, na.rm = TRUE)
  high_thresh <- quantile(x,p_upper, na.rm=TRUE)
  
  group_labs=c(paste("low: ", "(0% -", p_lower*100, "%)", sep = ""), paste("med: ", "(",p_lower*100,"% -", p_upper*100, "%)", sep = ""), paste("high: ", "(",p_upper*100,"% -100%)", sep = "") )
  
  grouped = ifelse(x <= low_thresh, group_labs[1], group_labs[2])
  grouped[which(x > high_thresh)] = group_labs[3]
  
  grouped = factor(grouped, levels=group_labs)
  return(grouped)
}
```


### 4. Subsetting by Project ID

```{r, echo=TRUE, warning=FALSE}

tcga_gbm <- subset(glioma_surv, project_id=='TCGA-GBM')

tcga_lgg <- subset(glioma_surv, project_id== 'TCGA-LGG')

```



## Survival models subset by Project ID {.tabset}

**There are 26 miclassified cases in the GBM project** 

```{r eval=TRUE}

table(tcga_gbm$WHO2021_recode)

```

**There are 97 misclassified cases in the LGG project** 

```{r}

table(tcga_lgg$WHO2021_recode)

```


```{r}


#diagnosis_data_frames <- lapply(unique_diagnoses, function(diagnosis) {
#  subset_gliomas <- subset(glioma_surv, WHO2021_recode == diagnosis)
#  return(subset_gliomas)
#})


gbm_df <- subset(glioma_surv, glioma_surv$WHO2021_recode == 'Glioblastoma, IDH-wt' )
astro_df <- subset(glioma_surv, glioma_surv$WHO2021_recode == 'Astrocytoma, IDH-mutant' )
oligo_df <- subset(glioma_surv, glioma_surv$WHO2021_recode == 'Oligodendroglioma, IDH-mutant' )
diff_glio_df <- subset(glioma_surv, glioma_surv$WHO2021_recode == 'Diffuse Glioma, IDH-mutant, NEC' )

#tcga_lgg <- subset(tcga_lgg, tcga_lgg$WHO2021_recode != 'Glioblastoma, IDH-wt' )

#retroelements <- c(names(glioma_counts[, num_columns]))

```

## Box Plots {.tabset}
### Differential counts of GBM (recoded) by project ID

```{r}
ggplot(gbm_df, aes(x = project_id, y = Alu, fill=project_id)) +
  geom_boxplot() +
  labs(title = "ALU Counts by project (within GBM)", x = "Project", y = "RNA Sequencing Counts")


```

```{r}

create_boxplot <- function(data, y_variable) {
  box_plot <- ggplot(data, aes(x = project_id, y = data[[y_variable]], fill = project_id)) +
    geom_boxplot() +
    labs(title = paste(toupper(y_variable), "Counts by project"),
         x = "Project",
         y = "RNA Sequencing Counts") 
  print(box_plot)
}

```

```{r}

for (re in names(num_columns)[1:49]){ #To exclude the IID column
  
  box_plot <- create_boxplot(gbm_df, re)
  
}

```


### Differential counts of Astros (recoded) by project ID

```{r}

for (re in names(num_columns)[1:49]){
  
  box_plot <- create_boxplot(astro_df, re)
  
}

```

### Differential counts of Oligos (recoded) by project ID

```{r}

for (re in names(num_columns)[1:49]){
  
  box_plot <- create_boxplot(astro_df, re)
  
}

```


### Differential counts of DIffuse glioma NEC (recoded) by project ID

```{r}

for (re in names(num_columns)[1:49]){
  
  box_plot <- create_boxplot(diff_glio_df, re)
  
}

```
