---
title: '2023-10 Survival by RE Family Adjusted by Project ID'
subtitle: RE project
output:
  html_document:
    df_print: paged
    number_sections: false
    toc: yes
    toc_depth: '3'
    toc_float: yes
    code_folding: hide
    theme: cosmo
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(survival)
library(survminer)
library(data.table)
library(table1)
library(reshape2)
library(doParallel)
library(foreach)
library(DESeq2)
library(limma)
library(edgeR)
library(kableExtra)
source("config.R")

```

## Data Cleanup {.tabset}

### 1. Normalizing counts

```{r normalization}
raw_counts <- readRDS(paste0(data_path,'/Data/RE_all_repFamily_1_raw_counts.RDS'))
normal_counts <- calcNormFactors(raw_counts, method = 'RLE')
normal_counts <- as.data.frame(cpm(normal_counts, normalized.lib.sizes=T))
```

### 2. Parsing through IDs

```{r combine-ids}
glioma_dt <- read.csv(paste0(data_path,"/Data/glioma_dt.csv"))
glioma_dt <- data.table(glioma_dt)
glioma_dt$case_submitter_id <- substr(glioma_dt$case_submitter_id, 6, 12)

```

### 3. Recoding gliomas


```{r glioma_reclass}

#GBMS
glioma_dt$WHO2021_recode <- ifelse(glioma_dt$IDH =='WT', 'Glioblastoma, IDH-wt', 'Diffuse Glioma, IDH-mutant, NEC')

#Oligos

glioma_dt <- glioma_dt %>%
  mutate(WHO2021_recode = ifelse(!is.na(x1p19q) & !is.na(IDH) & IDH == 'Mutant' & x1p19q == 'codel',
                                 'Oligodendroglioma, IDH-mutant', WHO2021_recode))

glioma_dt <- glioma_dt %>%
  mutate(WHO2021_recode = ifelse(!is.na(IDH) & IDH == 'Mutant' & TERT == 'Mutant' & WHO2021_recode == 'Diffuse Glioma, IDH-mutant, NEC',
                                 'Oligodendroglioma, IDH-mutant', WHO2021_recode))


#Astros
glioma_dt <- glioma_dt %>%
  mutate(WHO2021_recode = ifelse(!is.na(IDH) & !is.na(ATRX.status) & IDH == 'Mutant' & ATRX.status == 'Mutant',
                                 'Astrocytoma, IDH-mutant', WHO2021_recode))
```



```{r}

count_ids <- c(colnames(normal_counts))
count_ids <- substr(count_ids, 1, 7)

glioma_ids <- glioma_dt$case_submitter_id

glioma_counts <- normal_counts[, grepl(paste(glioma_ids, collapse ="|"), names(normal_counts))]
glioma_counts <- t(glioma_counts)
glioma_counts <- data.frame(glioma_counts)
glioma_counts$IID <- rownames(glioma_counts)
glioma_counts$IID <- rownames(glioma_counts)
rownames(glioma_counts) <- NULL
glioma_counts$IID <- substr(glioma_counts$IID, 1, 7)
rm(list =c('normal_counts', 'raw_counts'))
```



```{r}

glioma_surv <- merge(glioma_counts, glioma_dt, by.x="IID", by.y="case_submitter_id")
num_columns <- sapply(glioma_counts, is.numeric)

```

### Perecentile function

```{r percentile-function}
is_in_top_percentile <- function(x, p) {
  top_percentile <- quantile(x, p, na.rm = TRUE)
  ifelse(x >= top_percentile, 1, 0)
}


```

### Geno's tertile function

```{r}
# Function to group into low mid and high groups

percent_groups <- function(x, p_lower, p_upper) {
  
  low_thresh <- quantile(x, p_lower, na.rm = TRUE)
  high_thresh <- quantile(x,p_upper, na.rm=TRUE)
  
  group_labs=c(paste("low: ", "(0% -", p_lower*100, "%)", sep = ""), paste("med: ", "(",p_lower*100,"% -", p_upper*100, "%)", sep = ""), paste("high: ", "(",p_upper*100,"% -100%)", sep = "") )
  
  grouped = ifelse(x <= low_thresh, group_labs[1], group_labs[2])
  grouped[which(x > high_thresh)] = group_labs[3]
  
  grouped = factor(grouped, levels=group_labs)
  return(grouped)
}
```


### 4. Subsetting by Project ID

```{r, echo=TRUE, warning=FALSE}

tcga_gbm <- subset(glioma_surv, project_id=='TCGA-GBM')

tcga_lgg <- subset(glioma_surv, project_id== 'TCGA-LGG')

```



## Survival models subset by Project ID {.tabset}

**There are 26 miclassified cases in the GBM project** 

```{r eval=TRUE}

table(tcga_gbm$WHO2021_recode)

```

**There are 97 misclassified cases in the LGG project** 

```{r}

table(tcga_lgg$WHO2021_recode)

```


```{r}

tcga_gbm <- subset(tcga_gbm, tcga_gbm$WHO2021_recode == 'Glioblastoma, IDH-wt' )

tcga_lgg <- subset(tcga_lgg, tcga_lgg$WHO2021_recode != 'Glioblastoma, IDH-wt' )

retroelements <- c(names(glioma_counts[, num_columns]))

```


```{r, echo=FALSE}
#deleting objects that will not be used to free memory
suppressWarnings({rm(list =c('glioma_subsets', 'glioma_surv', 'glioma_counts', 'subtype_list'))})
```

## Models {.tabset}

### TCGA Low Grade Glioma

```{r}
lgg_re_coefs <- data.frame(re = character(), coef= numeric(), exp_coef = numeric(), se_coef = numeric(), z= numeric(), p_value = numeric(), stringsAsFactors = FALSE)
lgg_model_list <- list()

suppressWarnings({for (re in retroelements) { 
  base_formula <- as.formula(paste0("Surv(Survival_months, Vital_status) ~ Age + sex + race + WHO2021_recode + ", re))
  surv_model <- coxph(base_formula, data = tcga_lgg)
  p_value <- summary(surv_model)$coefficients[40]
  
  
  if (!is.na(p_value) && p_value <= 0.051) {
        coef <- summary(surv_model)$coefficients[8]
        exp_coef <- summary(surv_model)$coefficients[16]
        se_coef <- summary(surv_model)$coefficients[24]
        z <- summary(surv_model)$coefficients[32]
        lgg_re_coefs <- rbind(lgg_re_coefs, data.frame(re = re, coef = coef, exp_coef = exp_coef, se_coef = se_coef, z = z, p_value = p_value))
        lgg_model_list[[re]] <- surv_model
 
  
  }
}
})

lgg_re_coefs <- lgg_re_coefs[order(lgg_re_coefs$p_value), ]

lgg_re_coefs

```


### TCGA-Glioblastoma

```{r}
gbm_re_coefs <- data.frame(re = character(), coef= numeric(), exp_coef = numeric(), se_coef = numeric(), z= numeric(), p_value = numeric(), stringsAsFactors = FALSE)
gbm_model_list <- list()

suppressWarnings({for (re in retroelements) { 
  base_formula <- as.formula(paste0("Surv(Survival_months, Vital_status) ~ Age + sex + race + ", re))
  surv_model <- coxph(base_formula, data = tcga_gbm)
  p_value <- summary(surv_model)$coefficients[30]
  
  if (!is.na(p_value) && p_value <= 0.051) {
        coef <- summary(surv_model)$coefficients[6]
        exp_coef <- summary(surv_model)$coefficients[12]
        se_coef <- summary(surv_model)$coefficients[18]
        z <- summary(surv_model)$coefficients[24]
        gbm_re_coefs <- rbind(gbm_re_coefs, 
                              data.frame(re = re, coef = coef, exp_coef = exp_coef, se_coef = se_coef, z = z, p_value = p_value))
        gbm_model_list[[re]] <- surv_model

  
  }
}
})

gbm_re_coefs <- gbm_re_coefs[order(gbm_re_coefs$p_value), ]

gbm_re_coefs

```



## Plots {.tabset}


### TCGA Low Grade Glioma Project

```{r tcga_lgg_plot , echo=FALSE, fig.width=8, fig.height = 6 }
lgg_plots_list <- list()
lgg_re_list <- lgg_re_coefs$re
  #df <- coefs$re_coefs
  
for (re in lgg_re_list) {
  #re <- re_list[i,1]
  tcga_lgg$tertile_count <- as.factor(percent_groups(tcga_lgg[re], 0.33, 0.66))
    
  surv_obj <- Surv(tcga_lgg$Survival_months, tcga_lgg$Vital_status)
    
  km_curve <- surv_fit(surv_obj ~ tertile_count, data = tcga_lgg)
  plot <- ggsurvplot(km_curve, data = tcga_lgg, 
                       pval = TRUE, conf.int = TRUE, 
                       font.x = 7,
                       risk.table=TRUE, risk.table.fontsize = 4, risk.table.height=0.3,
                       title = paste0("KM Curve for ", re))
    
  lgg_plots_list[[re]] <- plot
  print(re)
  print(xtabs( ~ WHO2021_recode + tertile_count, data=tcga_lgg))
  print(plot)
}

```

### TCGA GBM project
```{r tcga_gbm_plot , echo=FALSE, fig.width=8, fig.height = 6 }
gbm_plots_list <- list()
gbm_re_list <- gbm_re_coefs$re
  #df <- coefs$re_coefs
  
for (re in gbm_re_list) {
  #re <- re_list[i,1]
  tcga_gbm$tertile_count <- as.factor(percent_groups(tcga_gbm[re], 0.33, 0.66))
    
  surv_obj <- Surv(tcga_gbm$Survival_months, tcga_gbm$Vital_status)
    
  km_curve <- surv_fit(surv_obj ~ tertile_count, data = tcga_gbm)
  plot <- ggsurvplot(km_curve, data = tcga_gbm, 
                       pval = TRUE, conf.int = TRUE, 
                       font.x = 7,
                       risk.table=TRUE, risk.table.fontsize = 4, risk.table.height=0.3,
                       title = paste0("KM Curve for ", re))
    
  gbm_plots_list[[re]] <- plot
  print(re)
  print(xtabs( ~ WHO2021_recode + tertile_count, data=tcga_gbm))
  print(plot)
}

```


